---
title: "Take-home Exercise 2"
author: "Zhiyi Hao"
date: "2026/03/01"
format:
  html:
    backgroundcolor: white
    fontcolor: black
    linkcolor: blue
---

# 2.1 The Task

# 2.2 Getting Started

First we loads the necessary R packages and then reads the cleaned transaction dataset into R.

```{r}
# load required packages
pacman::p_load(readxl, tidyverse, ggplot2, plotly, DT, cluster)
# import dataset
df <- read_excel("data/data.xlsx")
```

# 

2.3 Data Preparation and Cleaning

Before performing analysis, the raw trade data must be correctly cleaned and formatted. This step includes checking for missing values and duplicate records, standardizing variable formats, and converting the dataset from a long format to a wide format.

```{r}
# check duplicate records
df %>%
  count(country, year, month, category) %>%
  filter(n > 1)
# clean variables
df_clean <- df %>%
  mutate(
    country = str_trim(str_to_lower(country)),
    category = str_trim(str_to_lower(category)),
    year = as.integer(year),
    month = as.integer(month),
    total = as.numeric(total)
  ) %>%
  filter(
    !is.na(country),
    !is.na(category),
    !is.na(year),
    !is.na(month),
    !is.na(total)
  )
#check data after cleaning
glimpse(df_clean)
head(df_clean)
summary(df_clean)
colSums(is.na(df_clean))
# reshape data from long to wide format
df_wide <- df_clean %>%
  pivot_wider(
    names_from = category,
    values_from = total
  ) %>%
  arrange(country, year, month)
#check reshaped data
glimpse(df_wide)
head(df_wide)
summary(df_wide)
# create derived trade indicators
df_wide <- df_wide %>%
  mutate(
    trade_balance = output - input,
    trade_ratio = if_else(input != 0, output / input, NA_real_)
  )
#final check
glimpse(df_wide)
head(df_wide)
summary(df_wide)

colSums(is.na(df_wide))
```

# 2.4 Cluster and Outlier Analysis Module

# 

## 2.4.1 Country Trade Behaviour Segmentation

Country-level features are created to summarise the overall trade behaviour of each country. The clustering result is then used to compare countries across different trade profile groups.

Step 1: Create country-level features

Monthly observations are aggregated into country-level summary features. These variables capture trade scale, balance, ratio, and variability.

```{r}
country_features <- df_wide %>%
  group_by(country) %>%
  summarise(
    avg_input = mean(input, na.rm = TRUE),
    avg_output = mean(output, na.rm = TRUE),
    avg_balance = mean(trade_balance, na.rm = TRUE),
    avg_ratio = mean(trade_ratio, na.rm = TRUE),
    sd_input = sd(input, na.rm = TRUE),
    sd_output = sd(output, na.rm = TRUE),
    .groups = "drop"
  )

```

Step 2: Check the feature table

The feature is to confirm that the variables are created correctly.

```{r}
glimpse(country_features)
head(country_features)
summary(country_features)
colSums(is.na(country_features))
```

Step 3: Prepare the clustering data

The country identifier is removed and the numeric variables are standardised before clustering.

```{r}
cluster_data <- country_features %>%
  select(where(is.numeric))

cluster_data_scaled <- scale(cluster_data)
```

Step 4: Run clustering

In this prototype, countries are segmented into three clusters using k-means clustering.

```{r}
set.seed(123)

km3 <- kmeans(cluster_data_scaled, centers = 3, nstart = 25)

country_features$cluster <- as.factor(km3$cluster)

head(country_features)
```

Step 5: Append cluster labels and inspect results

The cluster labels are appended back to the country-level feature table for interpretation and plotting.

```{r}
head(country_features)
summary(country_features)
```

Step 6:Visualising results

The first chart shows the number of countries in each cluster. This provides an overview of how countries are distributed across the selected groups.

```{r}
plot_table1 <- country_features %>%
  count(cluster)

p1 <- ggplot(plot_table1, aes(x = cluster, y = n, fill = cluster)) +
  geom_col() +
  labs(
    title = "Number of Countries in Each Cluster",
    x = "Cluster",
    y = "Count"
  ) +
  theme_minimal()

ggplotly(p1)
```

The second chart compares the main trade characteristics across clusters. This helps users understand how the groups differ in terms of average trade behaviour.

```{r}
plot_table2 <- country_features %>%
  group_by(cluster) %>%
  summarise(
    avg_input = mean(avg_input, na.rm = TRUE),
    avg_output = mean(avg_output, na.rm = TRUE),
    avg_balance = mean(avg_balance, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = c(avg_input, avg_output, avg_balance),
    names_to = "variable",
    values_to = "value"
  )

p2 <- ggplot(plot_table2, aes(x = variable, y = value, fill = cluster)) +
  geom_col(position = "dodge") +
  labs(
    title = "Average Input, Output and Balance by Cluster",
    x = "Variable",
    y = "Value"
  ) +
  theme_minimal()

ggplotly(p2)
```

```         
```

```{r}
plot_table3 <- country_features %>%
  group_by(cluster) %>%
  summarise(
    avg_ratio = mean(avg_ratio, na.rm = TRUE),
    .groups = "drop"
  )

p3 <- ggplot(plot_table3, aes(x = cluster, y = avg_ratio, fill = cluster)) +
  geom_col() +
  labs(
    title = "Average Trade Ratio by Cluster",
    x = "Cluster",
    y = "Average Trade Ratio"
  ) +
  theme_minimal()

ggplotly(p3)
```

```{r}
country_features %>%
  select(country, avg_input, avg_output, avg_ratio, cluster) %>%
  arrange(desc(avg_ratio))
```

Step 8: Interactive plot

```{r}
country_table <- country_features %>%
  mutate(
    avg_input = round(avg_input, 2),
    avg_output = round(avg_output, 2),
    avg_balance = round(avg_balance, 2),
    avg_ratio = round(avg_ratio, 2)
  ) %>%
  select(country, avg_input, avg_output, avg_balance, avg_ratio, cluster)

datatable(country_table)
```

Step 9: Results table

```{r}
datatable(country_features)
```

## 2.4.2 UI Design – Part 1

## 

## 2.4.3 Seasonal Pattern Clustering

```{r}
df_clustered <- df_wide %>%
  left_join(
    country_features %>% select(country, cluster),
    by = "country"
  )
```

```{r}
glimpse(df_clustered)
head(df_clustered)
summary(df_clustered)
```
```{r}

```

```{r}
monthly_cluster_summary <- df_clustered %>%
  group_by(cluster, month) %>%
  summarise(
    avg_input = mean(input, na.rm = TRUE),
    avg_output = mean(output, na.rm = TRUE),
    avg_balance = mean(trade_balance, na.rm = TRUE),
    .groups = "drop"
  )

```


```{r}
head(monthly_cluster_summary)
summary(monthly_cluster_summary)
```

```{r}
p3 <- ggplot(
  monthly_cluster_summary,
  aes(x = month, y = avg_output, color = cluster, group = cluster)
) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title = "Monthly Output Pattern by Cluster",
    x = "Month",
    y = "Average Output"
  ) +
  theme_minimal()

ggplotly(p3)
```

```{r}
p4 <- ggplot(
  monthly_cluster_summary,
  aes(x = month, y = avg_input, color = cluster, group = cluster)
) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title = "Monthly Input Pattern by Cluster",
    x = "Month",
    y = "Average Input"
  ) +
  theme_minimal()

ggplotly(p4)
```
## 2.4.4 UI Design – Part 2

## 2.4.5 Cluster Transition Analysis

## 

## 2.4.6 UI Design – Part 3

## 

## 2.4.7 Outlier Detection

## 

## 2.4.8 UI Design – Part 4

# 2.5 References
