---
title: "Hands-on Exercise 1"
author: "Zhiyi Hao"
date: "2026/01/14"
format:
  html:
    backgroundcolor: white
    fontcolor: black
    linkcolor: blue
---

# Getting Started

## Install and launching R packages

The code chunk below uses p_load() of pacman package to check if tidyverse packages are installed in the computer. If they are, then they will be launched into R.

```{r}
pacman::p_load(tidyverse)
```

## Importing the data

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

# Plotting a simple bar chart

```{r}
ggplot(data = exam_data,
       aes(x = RACE)) + 
  geom_bar()
```

# Plotting histogram and use of aes()

This code chunk is to change color of the histogram by aes()

```{r}
ggplot(data=exam_data, aes(x= MATHS, fill = GENDER)) + geom_histogram(bins=15, color="skyblue")
```

# Plotting density plot & boxplot

This code chunk is to plot two density lines by using aes()

```{r}
ggplot(data = exam_data, aes(x = MATHS, colour = GENDER)) + geom_density() + scale_colour_manual(values = c( "Female" = "red", "Male" = "skyblue" ))
```
