---
title: "Hands-on Exercise 1"
author: "Zhiyi Hao"
date: "2026/01/14"
format:
  html:
    backgroundcolor: white
    fontcolor: black
    linkcolor: blue
---

# Getting Started

## Install and launching R packages

The code chunk below uses p_load() of pacman package to check if tidyverse packages are installed in the computer. If they are, then they will be launched into R.

```{r}
pacman::p_load(tidyverse)
```

## Importing the data

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

# Plotting a simple bar chart

```{r}
ggplot(data = exam_data,
       aes(x = RACE)) + 
  geom_bar()
```

# Plotting Geometric Objects: geom_dotplot

This code chunk is used to plot a dot plot.

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot(dotsize = 0.5)
```

# Modifying a geometric object by changing geom()

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20,            
                 color="black",      
                 fill="light blue") 
```

# Plotting histogram and use of aes()

This code chunk is to change color of the histogram by aes()

```{r}
ggplot(data=exam_data, aes(x= MATHS, fill = GENDER)) + geom_histogram(bins=15, color="skyblue")
```

# Plotting density plot & boxplot

This code chunk is to plot two density lines by using aes()

```{r}
ggplot(data = exam_data, aes(x = MATHS, colour = GENDER)) + geom_density() + scale_colour_manual(values = c( "Female" = "red", "Male" = "skyblue" ))
```

# Plotting Geometric Objects: geom_violin

The code plots the distribution of Maths score by gender in violin plot by using geom_violin()

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_violin()
```

# Plotting scatterplot

The code chunk below plots a scatterplot showing the Maths and English grades of pupils by using geom_point()

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point()
```

# **Working with `stat()`**

add a stat\_() function and override the default geom

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot()
```

# Adding best curve for scatterplot

In the code chunk geom_smooth() is used to plot a best fit curve on the scatterplot

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(size=0.5)
```

The default smoothing method can be overridden as shown below.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              linewidth=0.5)
```

# Working with facet_wrap()

The code chunk plots a trellis plot using facet-wrap()

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_wrap(~ CLASS)
```

# facet_grid() function

The code chunk below plots a trellis plot using facet-wrap() The code chunk plots a trellis plot using facet_grid()

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_grid(~ CLASS)
```

# Working with Coordinate

The code chunk flips the horizontal bar chart into vertical bar chart by using coord_flip()

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip()
```

# Working with theme

The code chunk plot a horizontal bar chart using theme_gray()

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_gray()
```

A horizontal bar chart plotted using theme_classic().

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_classic()
```

A horizontal bar chart plotted using theme_minimal().

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_minimal()
```
