---
title: "Hands-on Exercise 10"
author: "Zhiyi Hao"
date: "2026/03/18"
format:
  html:
    backgroundcolor: white
    fontcolor: black
    linkcolor: blue
---

This hands-on Exercise 10 we will go through chapter 31 for studying visualizing and analysing geographic data.

# **31 Information Dashboard Design: R methods**

## **31.1 Overview**

In this hands-on exercise, i am going to be able to:

-   create bullet chart by using ggplot2,

-   create sparklines by using ggplot2 ,

-   build industry standard dashboard by using R Shiny.

## **31.2 Getting started**

For the purpose of this hands-on exercise, the following R packages will be used.

```{r}
pacman::p_load(lubridate, ggthemes, reactable,
reactablefmtr, gt, gtExtras, tidyverse)
```

-   [tidyverse](https://www.tidyverse.org/) provides a collection of functions for performing data science task such as importing, tidying, wrangling data and visualising data. It is not a single package but a collection of modern R packages including but not limited to [readr](https://readr.tidyverse.org/), [tidyr](https://tidyr.tidyverse.org/), [dplyr](https://dplyr.tidyverse.org/), [ggplot](https://ggplot2.tidyverse.org/), [tibble](https://tibble.tidyverse.org/), [stringr](https://stringr.tidyverse.org/), [forcats](https://forcats.tidyverse.org/) and [purrr](https://purrr.tidyverse.org/).

-   [lubridate](https://r4va.netlify.app/chap31) provides functions to work with dates and times more efficiently.

-   [ggthemes](https://r4va.netlify.app/chap31) is an extension of ggplot2. It provides additional [themes](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/) beyond the basic themes of ggplot2.

-   [gtExtras](https://jthomasmock.github.io/gtExtras/index.html) provides some additional helper functions to assist in creating beautiful tables with [gt](https://gt.rstudio.com/), an R package specially designed for anyone to make wonderful-looking tables using the R programming language.

-   [reactable](https://glin.github.io/reactable/index.html) provides functions to create interactive data tables for R, based on the [React Table](https://github.com/tanstack/table/tree/v7) library and made with [reactR](https://github.com/react-R/reactR).

-   [reactablefmtr](https://kcuilla.github.io/reactablefmtr/index.html) provides various features to streamline and enhance the styling of interactive reactable tables with easy-to-use and highly-customizable functions and themes.

## **31.3 Importing Microsoft Access database**

### **31.3.1 The data set**

For the purpose of this study, a personal database in Microsoft Access mdb format called *Coffee Chain* will be used. \### **31.3.1 Import** Cause I am mac user, i think couldn't work on the mdb, so I will start with rds.

### **31.3.3 Data Preparation**

The code chunk below is used to import *CoffeeChain.rds* into R.

```{r}
coffeechain <- read_rds("data/CoffeeChain.rds")
```

Note: This step is optional if *coffeechain* is already available in R.

The code chunk below is used to aggregate Sales and Budgeted Sales at the Product level.

```{r}
product <- coffeechain %>%
  group_by(`Product`) %>%
  summarise(`target` = sum(`Budget Sales`),
            `current` = sum(`Sales`)) %>%
  ungroup()
```

### **31.3.4 Bullet chart in ggplot2**

The code chunk below is used to plot the bullet charts using ggplot2 functions.

```{r}
ggplot(product, aes(Product, current)) + 
  geom_col(aes(Product, max(target) * 1.01),
           fill="grey85", width=0.85) +
  geom_col(aes(Product, target * 0.75),
           fill="grey60", width=0.85) +
  geom_col(aes(Product, target * 0.5),
           fill="grey50", width=0.85) +
  geom_col(aes(Product, current), 
           width=0.35,
           fill = "black") + 
  geom_errorbar(aes(y = target,
                    x = Product, 
                    ymin = target,
                    ymax= target), 
                width = .4,
                colour = "red",
                size = 1) +
  coord_flip()
```

## **31.4 Plotting sparklines using ggplot2**

In this section, we will learn how to plot sparklines by using ggplot2.

### **31.4.1 Preparing the data**

```{r}
sales_report <- coffeechain %>%
  filter(Date >= "2013-01-01") %>%
  mutate(Month = month(Date)) %>%
  group_by(Month, Product) %>%
  summarise(Sales = sum(Sales)) %>%
  ungroup() %>%
  select(Month, Product, Sales)
```

```         
```

The code chunk below is used to compute the minimum, maximum and end othe the month sales.

```{r}
mins <- group_by(sales_report, Product) %>% 
  slice(which.min(Sales))
maxs <- group_by(sales_report, Product) %>% 
  slice(which.max(Sales))
ends <- group_by(sales_report, Product) %>% 
  filter(Month == max(Month))
```

The code chunk below is used to compute the 25 and 75 quantiles.

```{r}
quarts <- sales_report %>% group_by(Product) %>% summarise(quart1 = quantile(Sales, 0.25), quart2 = quantile(Sales, 0.75)) %>% right_join(sales_report)
```

### **31.4.2 sparklines in ggplot2**

The code chunk used.

```{r}
ggplot(sales_report, aes(x=Month, y=Sales)) + 
  facet_grid(Product ~ ., scales = "free_y") + 
  geom_ribbon(data = quarts, aes(ymin = quart1, max = quart2), 
              fill = 'grey90') +
  geom_line(size=0.3) +
  geom_point(data = mins, col = 'red') +
  geom_point(data = maxs, col = 'blue') +
  geom_text(data = mins, aes(label = Sales), vjust = -1) +
  geom_text(data = maxs, aes(label = Sales), vjust = 2.5) +
  geom_text(data = ends, aes(label = Sales), hjust = 0, nudge_x = 0.5) +
  geom_text(data = ends, aes(label = Product), hjust = 0, nudge_x = 1.0) +
  expand_limits(x = max(sales_report$Month) + 
                  (0.25 * (max(sales_report$Month) - min(sales_report$Month)))) +
  scale_x_continuous(breaks = seq(1, 12, 1)) +
  scale_y_continuous(expand = c(0.1, 0)) +
  theme_tufte(base_size = 3, base_family = "Helvetica") +
  theme(axis.title=element_blank(), axis.text.y = element_blank(), 
        axis.ticks = element_blank(), strip.text = element_blank())
```

## **31.5 Static Information Dashboard Design: gt and gtExtras methods**

By visiting the websites, I have found these two observations:

-   The **gt** package provides a flexible and elegant framework for building high-quality tables in R, with strong support for formatting and presentation.

-   The **gtExtras** package extends these capabilities by offering convenient helper functions and additional visual features, making it easier to create more polished and dashboard-style tabular displays.

### **31.5.1 Plotting a simple bullet chart**

In this section, you will learn how to prepare a bullet chart report by using functions of gt and gtExtras packages.

```{r}
product %>% gt::gt() %>% gt_plt_bullet(column = current, target = target, width = 60, palette = c("lightblue", "black")) %>% gt_theme_538()
```

## **31.6 sparklines: gtExtras method**

Before we can prepare the sales report by product by using gtExtras functions, code chunk below will be used to prepare the data.

```{r}
report <- coffeechain %>%
  mutate(Year = year(Date)) %>%
  filter(Year == "2013") %>%
  mutate (Month = month(Date, 
                        label = TRUE, 
                        abbr = TRUE)) %>%
  group_by(Product, Month) %>%
  summarise(Sales = sum(Sales)) %>%
  ungroup()
```

<div>

Note:

It is important to note that one of the requirement of gtExtras functions is that almost exclusively they require you to pass data.frame with list columns. In view of this, code chunk below will be used to convert the report data.frame into list columns.

</div>

```{r}
report %>% group_by(Product) %>% summarize('Monthly Sales' = list(Sales), .groups = "drop")
```

### **31.6.1 Plotting Coffechain Sales report**

```{r}
report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales), 
            .groups = "drop") %>%
   gt() %>%
   gt_plt_sparkline('Monthly Sales',
                    same_limit = FALSE)
```

### **31.6.2 Adding statistics**

First, calculate summary statistics by using the code chunk below.

```{r}
report %>% 
  group_by(Product) %>% 
  summarise("Min" = min(Sales, na.rm = T),
            "Max" = max(Sales, na.rm = T),
            "Average" = mean(Sales, na.rm = T)
            ) %>%
  gt() %>%
  fmt_number(columns = 4,
    decimals = 2)
```

### **31.6.3 Combining the data.frame**

Next, use the code chunk below to add the statistics on the table.

```{r}
spark <- report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales), 
            .groups = "drop")
```

```{r}
sales <- report %>% 
  group_by(Product) %>% 
  summarise("Min" = min(Sales, na.rm = T),
            "Max" = max(Sales, na.rm = T),
            "Average" = mean(Sales, na.rm = T)
            )
```

```{r}
sales_data = left_join(sales, spark)
```

### **31.6.4 Plotting the updated data.table**

```{r}
sales_data %>%
  gt() %>%
  gt_plt_sparkline('Monthly Sales',
                   same_limit = FALSE)
```

### **31.6.5 Combining bullet chart and sparklines**

Similarly, we can combining the bullet chart and sparklines using the steps below.

```{r}
bullet <- coffeechain %>%
  filter(Date >= "2013-01-01") %>%
  group_by(`Product`) %>%
  summarise(`Target` = sum(`Budget Sales`),
            `Actual` = sum(`Sales`)) %>%
  ungroup() 
```

```{r}
sales_data = sales_data %>%
  left_join(bullet)
```

```{r}
sales_data %>%
  gt() %>%
  gt_plt_sparkline('Monthly Sales') %>%
  gt_plt_bullet(column = Actual, 
                target = Target, 
                width = 28,
                palette = c("lightblue", 
                          "black")) %>%
  gt_theme_538()
```

**Summary**

From a dashboard design perspective, reactable and reactablefmtr complement each other well. Reactable improves user interaction with features such as sorting, filtering, and pagination, whereas reactablefmtr enhances the visual presentation of the table through compact and informative formatting elements. As a result, the table becomes more than a simple data display, and serves as an effective dashboard component for both exploration and communication.

## **31.7 Interactive Information Dashboard Design: reactable and reactablefmtr methods**

Before started, i get knowledge of these two packages works.

-   The **reactable** package is a flexible R package for building interactive tables with features such as sorting, filtering, pagination, and custom rendering.

-   The **reactablefmtr** package enhances **reactable** by adding convenient styling and formatting tools, including color scales, data bars, icons, and sparklines. Together, they are useful for creating interactive and visually appealing dashboard tables.

```{r}
remotes::install_github("timelyportfolio/dataui")
```

Then load the package onto R environment by using the code chunk below.

```{r}
library(dataui)
```

### **31.7.1 Plotting interactive sparklines**

Similar to gtExtras, to plot an interactive sparklines by using reactablefmtr package we need to prepare the list field by using the code chunk below.

```{r}
report <- report %>%
  group_by(Product) %>%
  summarize(`Monthly Sales` = list(Sales))
```

Next, [react_sparkline](https://kcuilla.github.io/reactablefmtr/reference/react_sparkline.html) will be to plot the sparklines as shown below.

Mac user can use this to first install remotes, then github one. The interactive table makes the dashboard more user-friendly by supporting quick exploration and comparison across products.

```{r}
# install.packages("remotes")
# remotes::install_github("timelyportfolio/dataui")
# install.packages("reactablefmtr")

# library(reactablefmtr)
# library(dataui)

library(reactablefmtr)
reactable(
  report,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(report)
    )
  )
)
```

### **31.7.2 Changing the pagesize**

By default the pagesize is 10. In the code chunk below, arguments defaultPageSize is used to change the default setting.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(report)
    )
  )
)
```

### **31.7.3 Adding points and labels**

In the code chunk below `highlight_points` argument is used to show the minimum and maximum values points and `label` argument is used to label first and last values.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        labels = c("first", "last")
        )
    )
  )
)
```

Adding highlighted points and labels makes important values easier to notice, especially the starting, ending, minimum, and maximum sales.

### **31.7.4 Adding reference line**

In the code chunk below `statline` argument is used to show the mean line.Darjeeling

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        statline = "mean"
        )
    )
  )
)
```

The reference line provides useful context by showing the average level, so individual product trends can be interpreted more meaningfully.

### **31.7.5 Adding bandline**

Instead adding reference line, bandline can be added by using the `bandline` argument.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        line_width = 1,
        bandline = "innerquartiles",
        bandline_color = "green"
        )
    )
  )
)
```

The bandline further improves interpretation by showing the typical range of variation, which helps distinguish normal fluctuation from unusual values. \### **31.7.6 Changing from sparkline to sparkbar**

Instead of displaying the values as sparklines, we can display them as sparkbars as shiwn below.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkbar(
        report,
        highlight_bars = highlight_bars(
          min = "red", max = "blue"),
        bandline = "innerquartiles",
        statline = "mean")
    )
  )
)
```

Using sparkbars instead of sparklines gives a clearer sense of magnitude, which is helpful when value comparison is more important than trend shape.

## **31.8 Reference**

Kam, T.S. (2026).R for visual Analytics. <https://r4va.netlify.app/chap31>

Thanks you prof for your very patient and nice teaching this term, hope I have to chance to participate your another class again!
